Chasing \rainbows{} with the Habitable Worlds Observatory

### 2023-04-15 src/tex/ms.tex
[showyourwork] overleaf sync

% Authors are in alphabetical order:
\author[Sophia R. Vaughan et al.]{%
    Sophia R. Vaughan$^{\affiliationlink{1},}$\thanks{Correspondence:  \url{sophia.vaughan@physics.ox.ac.uk}}
    Kimberly Bott$^{\affiliationlink{2},\affiliationlink{28}}$,
    Sarah L. Casewell$^{\affiliationlink{3}}$,
    Nicolas B. Cowan$^{\affiliationlink{4}}$,
    David S. Doelman$^{\affiliationlink{5,19}}$,
    \newauthor 
    Timothy D. Gebhard$^{\affiliationlink{6},\affiliationlink{29}}$,
    Matthew Kenworthy$^{\affiliationlink{5}}$,
    Johan Mazoyer$^{\affiliationlink{7}}$,
    Maxwell A. Millar-Blanchaer$^{\affiliationlink{8}}$,
    \newauthor 
    Olivier Absil$^{16}$,
    Lisa Altinier$^{13}$,
    Pierre Baudoz$^{7}$,
    Ruslan Belikov$^{24}$,
    Alexis Bidot$^{9}$,
    Markus J. Bonse$^{29}$,
    \newauthor 
    Bernhard Brandl$^{5}$,
    Alexis Carlotti$^{9}$,
    Elodie Choquet$^{13}$,
    Niyati Desai$^{17}$,
    Kevin Fogarty$^{24}$,
    Jules Fowler$^{12}$,
    \newauthor
    Yann Gutierrez$^{7,14,22}$,
    Olivier Guyon$^{15,25,26,27}$,
    Sebastiaan Y. Haffert$^{15}$,
    Olivier Herscovici-Schiller$^{14}$, 
    \newauthor
    Adrien Hours$^{9}$,
    Roser Juanola-Parramon$^{20,21}$,
    Elina Kleisioti $^{5}$,
    Lorenzo König$^{16}$,
    Mariya Krasteva$^{23}$, 
    \newauthor
    Iva Laginja$^{7}$,
    Rico Landman$^{5}$,
    Lucie Leboulleux$^{9}$,
    David Mouillet$^{9}$,
    Mamadou N’Diaye$^{10}$,
    Emiel H. Por$^{18}$,
    \newauthor
    Laurent Pueyo$^{18}$,
    Frans Snik$^{5}$,
    Daphne M. Stam$^{30}$,
    Victor Trees$^{31,32}$,
    Dirk van Dam$^{5}$,
    Kyle van Gorkom$^{15}$,
    Maaike van Kooten$^{11}$ 
    \newauthor \\%
    Affiliations are listed at the end of the paper
    % Timmy: I checked MNRAS and it seems that this is how they handle very long author lists. See, for example:
    % https://academic.oup.com/mnras/article/498/2/2354/5900562
    % I've also added some LaTeX magic to create links for the affiliations; if y'all agree that it's a good idea I can extend it to all authors?
}
\date{Draft version: \today}

@@

\begin{abstract}
NASA recently announced the Habitable Worlds Observatory (HWO), a coronagraphic mission to detect rocky planets in habitable zones, establish their habitability, and search for biosignatures. 
%Numerous instrumental and observational challenges have still to be overcome to design a coronagraphic mission with that goal. After a research phase aiming at the detection of a few tens of exo-Earth in the habitable zone, the Habitable Worlds Observatory will use its spectroscopic and polarimetric capabilities to probe these planets' atmosphere and surface properties. 
Surface liquid water is central to the definition of planetary habitability.
Photometric and polarimetric phase variations can show the presence of specularly reflecting oceans on exoplanets, in addition to being sensitive to rainbows and other cloud scattering effects. 
Direct imaging missions are optimised to detect planets near quadrature, which may obscure some of the phases at which these features are strongest. 
The range of scattering phases accessible for an exoplanet can be limited by its orbital inclination or the coronagraph's inner working angle. 
Planets that orbit close to edge-on have accessible phase angles limited by obscuration by the coronagraph. 
We use the list of target stars for the Habitable Worlds Observatory to estimate the number of exo-Earths that could be searched for non-Lambertian scattering phenomena. 
We find that exo-Earths in systems listed in this Habitable Worlds Observatory catalog will have a detectable Rayleigh scattering peak. 
The glint signature at scattering phases of $\sim$50--70$^\circ$ would be accessible to the Habitable Worlds Observatory in $\sim$14 systems, and the rainbow of water clouds at phases of 140--160$^\circ$ would be accessible in $\sim$42 systems, assuming a 62\,mas inner working angle (\IWA; 3$\lambda/D$ for a 6-m telescope at 600nm). Increasing the \IWA to 2$\lambda/D$, increases the number of systems with detectable rainbows and glints by factors of roughly 2 and 3, respectively. The number of systems in which an exo-Earth could be searched for these scattering phenomena increases with orbital eccentricity and scales approximately inversely with \IWA.     \end{abstract}

@@

%As ocean glint is one of the few sources of specular reflection and is strongly polarized, it provides one of the best means to detect a signature of habitability in exoplanet phase curves. 
Ocean glint is from the specular (Fresnel) reflection of light off of smooth liquid surfaces at crescent phases. In the total flux the signal of the glint will be most prominent at small planetary phase (large scattering) angles. The maximum polarization occurs at the Brewster angle determined by the refractive indices at the interface; between air and liquid water this occurs at 53 degrees (a scattering angle of 127 degrees) \citep[see, e.g. ][]{2008Icar..195..927W}. It therefore varies slightly by wavelength and by sea and atmosphere composition. In general on a planet the glint will be strongest in red wavelengths where the Rayleigh contribution to both the phase curve and attenuation are minimized \citep{Zugger_2011} and, conveniently, where the latitude-albedo effect (a degeneracy between polar brightness and glint) is also minimized \citep{2012ApJ...752L...3C}.  The strengthened glint signature at longer wavelengths has been observed in Earthshine measurements \citep{Emde2017,sterzik2019, takahashi2021}.
The broadening of the glint feature in the presence of wind-driven waves is well-described by the Cox-Munk solution \citep{CoxMunk1954}, shifts in both polarized and unpolarized light, and broadens across scattering angles in polarized light \citep{kopparla2018, Zugger_2010, treesandstam2019, trees2022}.

@@

In unpolarized light the forward scattering and backscatter from clouds can dramatically change the shape of a phase curve from Lambertian. While the glory and rainbow are technically visible in any phase curve, their strongly polarized nature coupled with the suppression of signal from other effects at gibbous and full phases in polarized light make them far easier to characterize in polarized light (both in the degree of polarization and in full Stokes measurements) \citep{karalidi2011, stam2008, treesandstam2019}. The rainbow angle depends on the refractive index and therefore varies with wavelength and depending on the species. The feature being far more narrow than the glint allows more exact determination of the species. The droplet size distribution also effects the shape of the rainbow and clouds of species in a solid phase can diverge from the Mie scattering that produces these features greatly, producing other optical effects particularly at narrow scattering angles \citep{bailey2007, hansentravis1974, Karalidi2012ice}. In this paper we define our optimal angle for rainbow detection as the peak of the water droplet rainbow in visible wavelengths.  However, extending observations to angles closer to the star (smaller scattering angles, larger phase angles) would allow the full shape of the rainbow to be characterized and ensure the characterization of rainbows from other possible HZ condensates. The glory is produced when the droplet radii much smaller than the wavelength of light and can allow further characterization as their angle, intensity, and shape also vary with the complex refractive index but primarily occur within 20 degrees scattering angle very close to full phase. %The glory (in the backscattering direction) and diffraction peak (in the forward scattering direction) are distinct features in the total reflected flux, but not in the degree of polarization. In principle, the degree of polarization goes to zero in the forward and backward scattering directions. See e.g. Fig. 1 of \citet{stam2008}, Fig. 5 of \citet{trees2022}, Fig. 3 of \citet{karalidi2011} and \citet{hansentravis1974}.}}

>>>>

% Authors are in alphabetical order:
\author[Sophia R. Vaughan et al.]{%
    Sophia R. Vaughan$^{\affiliationlink{1},}$\thanks{Correspondence:  \url{sophia.vaughan@physics.ox.ac.uk}}
    Kimberly Bott$^{\affiliationlink{2},\affiliationlink{28}}$,
    Sarah L. Casewell$^{\affiliationlink{3}}$,
    Nicolas B. Cowan$^{\affiliationlink{4}}$,
    David S. Doelman$^{\affiliationlink{5,19}}$,
    \newauthor 
    Timothy D. Gebhard$^{\affiliationlink{6},\affiliationlink{29}}$,
    Matthew Kenworthy$^{\affiliationlink{5}}$,
    Johan Mazoyer$^{\affiliationlink{7}}$,
    Maxwell A. Millar-Blanchaer$^{\affiliationlink{8}}$,
    \newauthor 
    Olivier Absil$^{16}$,
    Lisa Altinier$^{13}$,
    Pierre Baudoz$^{7}$,
    Ruslan Belikov$^{24}$,
    Alexis Bidot$^{9}$,
    Markus J. Bonse$^{29}$,
    \newauthor 
    Bernhard Brandl$^{5}$,
    Alexis Carlotti$^{9}$,
    Elodie Choquet$^{13}$,
    Niyati Desai$^{17}$,
    Kevin Fogarty$^{24}$,
    Jules Fowler$^{12}$,
    \newauthor
    Yann Gutierrez$^{7,14,22}$,
    Olivier Guyon$^{15,25,26,27}$,
    Sebastiaan Y. Haffert$^{15}$,
    Olivier Herscovici-Schiller$^{14}$, 
    \newauthor
    Adrien Hours$^{9}$,
    Roser Juanola-Parramon$^{20,21}$,
    Elina Kleisioti $^{5}$,
    Lorenzo König$^{16}$,
    Mariya Krasteva$^{23}$, 
    \newauthor
    Iva Laginja$^{7}$,
    Rico Landman$^{5}$,
    Lucie Leboulleux$^{9}$,
    David Mouillet$^{9}$,
    Mamadou N’Diaye$^{10}$,
    Emiel H. Por$^{18}$,
    \newauthor
    Laurent Pueyo$^{18}$,
    Frans Snik$^{5}$,
    Daphne M. Stam$^{30}$,
    Victor Trees$^{31,32}$,
    Dirk van Dam$^{5}$,
    Kyle van Gorkom$^{15}$,
    \newauthor
    Maaike van Kooten$^{11}$ 
    \newauthor \\%
    Affiliations are listed at the end of the paper
    % Timmy: I checked MNRAS and it seems that this is how they handle very long author lists. See, for example:
    % https://academic.oup.com/mnras/article/498/2/2354/5900562
    % I've also added some LaTeX magic to create links for the affiliations; if y'all agree that it's a good idea I can extend it to all authors?
}
\date{Draft version: \today}

@@

\begin{abstract}
NASA recently announced the Habitable Worlds Observatory (HWO), a coronagraphic mission to detect rocky planets in habitable zones, establish their habitability, and search for biosignatures. 
%Numerous instrumental and observational challenges have still to be overcome to design a coronagraphic mission with that goal. After a research phase aiming at the detection of a few tens of exo-Earth in the habitable zone, the Habitable Worlds Observatory will use its spectroscopic and polarimetric capabilities to probe these planets' atmosphere and surface properties. 
Surface liquid water is central to the definition of planetary habitability.
Photometric and polarimetric phase variations can show the presence of specularly reflecting oceans on exoplanets, in addition to being sensitive to rainbows and other cloud scattering effects. 
Direct imaging missions are optimised to detect planets near quadrature, which may obscure some of the phases at which these features are strongest. 
The range of scattering phases accessible for an exoplanet can be limited by its orbital inclination or the coronagraph's inner working angle. 
Planets that orbit close to edge-on have accessible phase angles limited by obscuration by the coronagraph. 
We use the list of target stars for the Habitable Worlds Observatory to estimate the number of exo-Earths that could be searched for non-Lambertian scattering phenomena. 
We find that exo-Earths in systems listed in this Habitable Worlds Observatory catalog will have a detectable Rayleigh scattering peak. 
The glint signature at planetary phase angles of 
%$\sim$50--70$^\circ$ 
$\sim$110--130$^\circ$ would be accessible to the Habitable Worlds Observatory in $\sim$14 systems, and the rainbow of water clouds at phases of 140--160$^\circ$ would be accessible in $\sim$42 systems, assuming a 62\,mas inner working angle (\IWA; 3$\lambda/D$ for a 6-m telescope at 600nm). Increasing the \IWA to 2$\lambda/D$, increases the number of systems with detectable rainbows and glints by factors of roughly 2 and 3, respectively. The number of systems in which an exo-Earth could be searched for these scattering phenomena increases with orbital eccentricity and scales approximately inversely with \IWA.     \end{abstract}

@@

%As ocean glint is one of the few sources of specular reflection and is strongly polarized, it provides one of the best means to detect a signature of habitability in exoplanet phase curves. 
Ocean glint is from the specular (Fresnel) reflection of light off of smooth liquid surfaces at crescent phases. In the total flux the signal of the glint will be most prominent at small planetary phase (large scattering) angles. The maximum polarization occurs at the Brewster angle determined by the refractive indices at the interface; between air and liquid water this occurs at 53 degrees  \citep[a scattering angle of 127 degrees; see e.g. ][]{2008Icar..195..927W}. It therefore varies slightly by wavelength and by sea and atmosphere composition. In general on a planet the glint will be strongest in red wavelengths where the Rayleigh contribution to both the phase curve and attenuation are minimized \citep{Zugger_2011} and, conveniently, where the latitude-albedo effect (a degeneracy between polar brightness and glint) is also minimized \citep{2012ApJ...752L...3C}.  The strengthened glint signature at longer wavelengths has been observed in Earthshine measurements \citep{Emde2017,sterzik2019, takahashi2021}.
The broadening of the glint feature in the presence of wind-driven waves is well-described by the Cox-Munk solution \citep{CoxMunk1954}, shifts in both polarized and unpolarized light, and broadens across scattering angles in polarized light \citep{kopparla2018, Zugger_2010, treesandstam2019, trees2022}.

@@

In unpolarized light the forward scattering and backscatter from clouds can dramatically change the shape of a phase curve from Lambertian. While the glory and rainbow are technically visible in any phase curve, their strongly polarized nature coupled with the suppression of signal from other effects at gibbous and full phases in polarized light make them far easier to characterize in polarized light (both in the degree of polarization and in full Stokes measurements) \citep{karalidi2011, stam2008, treesandstam2019}. The rainbow angle depends on the refractive index and therefore varies with wavelength and depending on the species. The feature being far more narrow than the glint allows more exact determination of the species. The droplet size distribution also effects the shape of the rainbow and clouds of species in a solid phase can diverge from the Mie scattering that produces these features greatly, producing other optical effects particularly at narrow scattering angles \citep{bailey2007, hansentravis1974, Karalidi2012ice}. In this paper we define our optimal angle for rainbow detection as the peak of the water droplet rainbow in visible wavelengths.  However, extending observations to angles closer to the star (smaller scattering angles, i.e.\ larger phase angles) would allow the full shape of the rainbow to be characterized and ensure the characterization of rainbows from other possible HZ condensates. The glory is produced when the droplet radii much smaller than the wavelength of light and can allow further characterization as their angle, intensity, and shape also vary with the complex refractive index but primarily occur within 20 degrees scattering angle very close to full phase. %The glory (in the backscattering direction) and diffraction peak (in the forward scattering direction) are distinct features in the total reflected flux, but not in the degree of polarization. In principle, the degree of polarization goes to zero in the forward and backward scattering directions. See e.g. Fig. 1 of \citet{stam2008}, Fig. 5 of \citet{trees2022}, Fig. 3 of \citet{karalidi2011} and \citet{hansentravis1974}.}}
